Leadership across cultures and contexts: an exploratory analysis

### 2020-05-09 leadership_across_cultures_contexts.Rmd
Revisions to feature discussions.

The distribution of text records by group context and subsistence strategy suggests that although leaders emerge within similar social contexts across populations, it occurs at variable frequencies. See Figure \@ref(fig:grouptype). Leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, despite potential for residential mobility, do rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

One possible interpretation of the negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level -- more similar to leadership systems of WEIRD populations -- leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintains important leadership positions within the group and represents the group to governmental bodies @aa01-062 [p. 173] describes that in these local 'clubs',

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of the nonindustrial cultures sampled here.

@@

The Prosocial qualities cluster comprised subclusters we labeled, "Cultural conformity" and "Prosocial competencies", the later of which included a mix of positive personal traits, such as generosity and proper behavior, as well as interpersonal skills, having a favorable personality, oratory skill, and charisma, which generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The larger Prosocial qualities cluster also corresponds to @henrich_big_2015's Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership. A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity).

The three most frequent leader quality variables were found in the Individual qualities cluster, which had two smaller and strongly supported subclusters: "Social, reproductive, and material success", which included traits indicative of the benefits of successful leadership, and a more general "Competencies" cluster which included qualities that are not necessarily prosocial. Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Again, some variables in this subcluster, such as *Ingroup member* and *Loyalty*, did not quite fit the theme. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

The Competencies subcluster contained a grouping of variables associated with dominance-based leadership, including *Coercive authority* and *Aggressiveness* [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a grouping that included *Knowledgeable/intelligent* and *Experienced/accomplished*, variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), a point to which we shall return. Interestingly, *Feared* and *Supernatural* clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide a necessary and valued public good, have special expertise and abilities, are feared and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

@@

The leader function variables formed two major clusters, one that we identified as "Prosociality" and one that we identified as "Management". The prosocial cluster included *Distribute resources*, *Protection*, and *Prosocial investment*, which are functions that are commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain social influence via continued community support and the demands of followers can shape the functions of leaders to meet local needs.

The management cluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

Other variables in this subcluster indicated more generic group strategizing, such as *Strategic planning* and *Group representative* [@de_cremer_intergroup_2002]. In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services.

The larger organization subcluster included variables such as *Organize cooperation*, *Control economics*, and *Military command*. This subcluster aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:

>>>>

The distribution of text records by group context and subsistence strategy suggests that although leaders emerge within similar social contexts across populations, it occurs at variable frequencies. See Figure \@ref(fig:grouptype). Leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese hunter-gatherers by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, despite potential for residential mobility, do rely in part on a highly defensible resource base (i.e., livestock) and exhibit political leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992; @glowacki_evolutionary_2017]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

One possible interpretation of the negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among nonindustrial agricultural populations which are characterized by leadership systems beyond the community level -- more similar to leadership systems of WEIRD populations -- the importance of leadership within residential and family groups persists. For example, among rural Korean populations a chairwomen of local organizations maintains important, possibly informal, leadership positions within the group and represents the group to governmental bodies. @aa01-062 [p. 173] describes that in these local 'clubs',

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of the nonindustrial cultures sampled here. Clearly there is substantial diversity in many dimensions of leadership among non-WEIRD populations.

@@

<!-- DISCUSSES PROSOCIALITY FEATURE -->
The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural values shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to @henrich_big_2015's Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

<!-- DISCUSSES SUCCESS FEATURE -->
Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

<!-- DISCUSSES COMPETENCIES FEATURE -->
The individual-level competencies was associated with variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a grouping that included qualities of knowledgeable and intelligent and experienced and accomplished, traits that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), a point to which we shall return. Interestingly, being feared and having supernatural qualities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide a necessary and valued public good, have special expertise and abilities, are feared and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
The prosocial feature of leader functions included distributing resources, providing protection, and various prosocial investments, which are functions commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain social influence via continued community support and the demands of followers can shape the functions of leaders to meet local needs.

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closedly to what @van_vugt_cognitive_2007 define as *strategic* versus *coordinating* leadership.

Leaders, possibly due to their special capacities, are able to devise and impliment strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

The feature of leader organizatonal functions aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:
